Automated Assessment of Residual Plots with Computer Vision Models

### 2025-06-13 paper.Rmd
Reduce MVI section

To reduce computational cost, a precomputed lattice of $\hat{D}$ quantiles under $H_0$ for various sample sizes can be used. By matching observed values to the closest entries in this lattice, approximate $p$-values can be obtained efficiently, with minor loss of precision.

@@

While statistical testing is a powerful tool for detecting model violations, it can become cumbersome and time-consuming when quick decisions are needed, particularly due to the need to evaluate numerous null plots. In practice, a more convenient and immediate method for assessing model performance is often required. This is where an index, such as the Model Violations Index (MVI), becomes valuable. It offers a straightforward way to quantify deviations from model assumptions, enabling rapid assessment and easier comparison across models.

The estimator $\hat{D}$ measures the difference between the true residual distribution and the reference residual distribution, a difference primarily arises from deviations in model assumptions. The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violations index (MVI).

Note that if more observations are used for estimating the linear regression, the result of Equation \ref{eq:kl-1} will increase, as the integration will be performed with larger $n$. For a given data generating process, $D$ typically increases logarithmically with the number of observations. This behavior comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.

Since $\hat{D}$ is an estimate of $D$, it is expected that a larger number of observations will also lead to a higher $\hat{D}$. However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations. In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.

Therefore, the Model Violations Index (MVI) can be proposed as

@@

\noindent where $C$ is a large enough constant keeping the result positive and the term $-\log(n)$ is used to offset the increase in $D$ due to sample size.

@@

The architecture of the computer vision model is adapted from the well-established VGG16 architecture [@simonyan2014very]. While more recent architectures like ResNet [@he2016deep] and DenseNet[@huang2017densely], have achieved even greater performance, VGG16 remains a solid choice for many applications due to its simplicity and effectiveness. Our decision to use VGG16 aligns with our goal of starting with a proven and straightforward model. Figure \ref{fig:cnn-diag} provides a diagram of the architecture. More details about the neural network layers used in this study are provided in the Appendix.

>>>>

To reduce computational cost, a pre-computed lattice of $\hat{D}$ quantiles under $H_0$ for various sample sizes can be used. By matching observed values to the closest entries in this lattice, approximate $p$-values can be obtained efficiently, with minor loss of precision.

@@

In \ref{sec-model-lineup-evaluation}, we noted that a pre-computed lattice of $\hat{D}$ quantiles can reduce the computation cost of lineup tests. Another practical approach is to assess model performance directly using the value of $\hat{D}$.

The estimator $\hat{D}$ captures the difference between the true and reference residual distributions, which reflects the extent of model violations, making it instrumental in forming a model violations index (MVI). However, when more observations are used in regression, the value of $\hat{D}$ tends to increase logarithmically. This is because $D = \log(1 + D_{KL})$, and under the assumption of independence, $D_{KL}$ is the sum of $D_{KL}^{(i)}$ across all observations. This does not mean that $\hat{D}$ becomes less reliable. In fact, larger samples often make model violations more visible in residual plots, unless strong overlapping masks the patterns.

However, to create a standardized and generalizable index, it is important to adjust for the effect of sample size. Therefore, the Model Violations Index (MVI) is proposed as

@@

\noindent where $C$ is a sufficiently large constant to ensure the result remains positive, and the $\log(n)$ term offset the increase in $D$ with larger sample sizes.

@@

The architecture of the computer vision model $f_{CV}$ is adapted from the well-established VGG16 architecture [@simonyan2014very]. While more recent architectures like ResNet [@he2016deep] and DenseNet[@huang2017densely], have achieved even greater performance, VGG16 remains a solid choice for many applications due to its simplicity and effectiveness. Our decision to use VGG16 aligns with our goal of starting with a proven and straightforward model. Figure \ref{fig:cnn-diag} provides a diagram of the architecture. More details about the neural network layers used in this study are provided in the Appendix.

### 2025-06-13 paper.tex
Reduce MVI section

To reduce computational cost, a precomputed lattice of \(\hat{D}\)
quantiles under \(H_0\) for various sample sizes can be used. By
matching observed values to the closest entries in this lattice,
approximate \(p\)-values can be obtained efficiently, with minor loss of
precision.

@@

While statistical testing is a powerful tool for detecting model
violations, it can become cumbersome and time-consuming when quick
decisions are needed, particularly due to the need to evaluate numerous
null plots. In practice, a more convenient and immediate method for
assessing model performance is often required. This is where an index,
such as the Model Violations Index (MVI), becomes valuable. It offers a
straightforward way to quantify deviations from model assumptions,
enabling rapid assessment and easier comparison across models.

The estimator \(\hat{D}\) measures the difference between the true
residual distribution and the reference residual distribution, a
difference primarily arises from deviations in model assumptions. The
magnitude of \(D\) directly reflects the degree of these deviations,
thus making \(\hat{D}\) instrumental in forming a model violations index
(MVI).

Note that if more observations are used for estimating the linear
regression, the result of Equation \ref{eq:kl-1} will increase, as the
integration will be performed with larger \(n\). For a given data
generating process, \(D\) typically increases logarithmically with the
number of observations. This behavior comes from the relationship
\(D = \text{log}(1 + D_{KL})\), where
\(D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}\) under the assumption of
independence.

Since \(\hat{D}\) is an estimate of \(D\), it is expected that a larger
number of observations will also lead to a higher \(\hat{D}\). However,
this does not imply that \(\hat{D}\) fails to accurately represent the
extent of model violations. In fact, when examining residual plots with
more observations, we often observe a stronger visual signal strength,
as the underlying patterns are more likely to be revealed, except in
cases of significant overlapping.

Therefore, the Model Violations Index (MVI) can be proposed as

@@

\noindent where \(C\) is a large enough constant keeping the result
positive and the term \(-\log(n)\) is used to offset the increase in
\(D\) due to sample size.

@@

The architecture of the computer vision model is adapted from the
well-established VGG16 architecture \citep{simonyan2014very}. While more
recent architectures like ResNet \citep{he2016deep} and
DenseNet\citep{huang2017densely}, have achieved even greater
performance, VGG16 remains a solid choice for many applications due to
its simplicity and effectiveness. Our decision to use VGG16 aligns with
our goal of starting with a proven and straightforward model. Figure
\ref{fig:cnn-diag} provides a diagram of the architecture. More details
about the neural network layers used in this study are provided in the
Appendix.

>>>>

To reduce computational cost, a pre-computed lattice of \(\hat{D}\)
quantiles under \(H_0\) for various sample sizes can be used. By
matching observed values to the closest entries in this lattice,
approximate \(p\)-values can be obtained efficiently, with minor loss of
precision.

@@

In \ref{sec-model-lineup-evaluation}, we noted that a pre-computed
lattice of \(\hat{D}\) quantiles can reduce the computation cost of
lineup tests. Another practical approach is to assess model performance
directly using the value of \(\hat{D}\).

The estimator \(\hat{D}\) captures the difference between the true and
reference residual distributions, which reflects the extent of model
violations, making it instrumental in forming a model violations index
(MVI). However, when more observations are used in regression, the value
of \(\hat{D}\) tends to increase logarithmically. This is because
\(D = \log(1 + D_{KL})\), and under the assumption of independence,
\(D_{KL}\) is the sum of \(D_{KL}^{(i)}\) across all observations. This
does not mean that \(\hat{D}\) becomes less reliable. In fact, larger
samples often make model violations more visible in residual plots,
unless strong overlapping masks the patterns.

However, to create a standardized and generalizable index, it is
important to adjust for the effect of sample size. Therefore, the Model
Violations Index (MVI) is proposed as

@@

\noindent where \(C\) is a sufficiently large constant to ensure the
result remains positive, and the \(\log(n)\) term offset the increase in
\(D\) with larger sample sizes.

@@

The architecture of the computer vision model \(f_{CV}\) is adapted from
the well-established VGG16 architecture \citep{simonyan2014very}. While
more recent architectures like ResNet \citep{he2016deep} and
DenseNet\citep{huang2017densely}, have achieved even greater
performance, VGG16 remains a solid choice for many applications due to
its simplicity and effectiveness. Our decision to use VGG16 aligns with
our goal of starting with a proven and straightforward model. Figure
\ref{fig:cnn-diag} provides a diagram of the architecture. More details
about the neural network layers used in this study are provided in the
Appendix.
